---
title: "Discrete event Systems simulation - León XIII Clinic - Emergency department"
author: "Daniel"
date: "21/2/2020"
output: 
  html_document: 
    df_print: kable
    number_sections: yes
    theme: cosmo
    toc: yes
    toc_float: true
    code_folding: "hide"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This project aims to fit distributions to the services inside the emergency department of the León XIII clinic, located in Medellín - Colombia using real data from the users and their arrivals. Extensive use of the `fitdistrplus` package is made in the development of this methodology.

After the distributions are fitted, discrete event simulation will be performed using Simul8 to stablish performance measures and prescriptive analytics that could improve the decision making process. 

# Libraries

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(fitdistrplus)
library(lubridate)
library(ggthemes)
```

# Data reading

Admission office dynamics could be hidden inside the data repetition, need to assess whether including the triages makes any difference whatsoever.

```{r}
datos_complementarios <- read_csv2("datasets/tiempos_ingresos_15k.csv", locale = readr::locale(encoding = "latin1")) %>% 
  mutate(servicio = as_factor(servicio),
         servicio_detalle = as_factor(servicio_detalle),
         nivelPrioridad = as_factor(nivelPrioridad)) %>% 
  mutate_at(vars(salida_taquilla, entrada_triage, fin_triage, entrada_servicio, salida_servicio), as_datetime)

raw_data <- read_csv("datasets/tiempos_ingresos_10k.csv") %>% 
  inner_join(datos_complementarios %>%
               distinct(atencion, .keep_all = TRUE) %>% 
              dplyr::select(atencion, nivelPrioridad))
head(raw_data)
```

These two datasets are joined because in the first extraction of raw_data performed with the IT department, the triage type information was not on the initial SQL query, therefore a later query called `datos_complementarios` was made with the triage information on it, that was bigger than the original raw_data and included the triage information and more entries. The original data remained unaltered and the triage information was joined to it (10000 observations).

## Study Interval

```{r}
raw_data %>%
  summarize(min(entrada_triage))

raw_data %>%
  filter(servicio == "Urgencias" | servicio == "Urgencias Pediatria") %>% 
  ggplot(aes(x = entrada_triage)) +
  geom_histogram(bins = 44, color = "black", fill = "skyblue") +
  labs(title = "Intervalo del estudio realizado por fecha",
       x = "Fecha del registro")
```

Interval: 2020-01-07 12:51:12 to 2020-02-21 17:00:00

## Hourly distribution by service

```{r}
raw_data %>%
  ggplot(aes(x = hour(entrada_triage), fill = servicio)) +
  geom_histogram(bins = 24, color = "black") +
  labs(title = "Horas más concurridas por servicio")
```

Interval: 2020-01-07 12:51:12 to 2020-02-21 17:00:00

## daily distribution by service

```{r}
raw_data %>%
  ggplot(aes(x = day(entrada_triage), fill = servicio)) +
  geom_histogram(bins = 31, color = "black") +
  labs(title = "llegadas por dias del mes por servicio")
```

Interval: 2020-01-07 12:51:12 to 2020-02-21 17:00:00

## weekday distribution by service

```{r}
raw_data %>%
  ggplot(aes(x = weekdays(entrada_triage), fill = servicio)) +
  geom_bar( color = "black") +
  labs(title = "llegadas por dia de la semana por servicio")
```

# Creation of the duration of events

```{r}
clean_data <- raw_data %>%
  distinct(salida_taquilla, .keep_all = TRUE)

tiempo_llegadas <- diff(sort(clean_data$salida_taquilla))

tiempos <- clean_data %>% 
  mutate(cola_triage = entrada_triage - salida_taquilla,
         dur_triage = fin_triage - entrada_triage,
         cola_servicio = entrada_servicio - fin_triage,
         dur_servicio = salida_servicio - entrada_servicio) %>%
  mutate_at(vars(cola_triage, dur_triage, cola_servicio, dur_servicio), as.numeric) %>% 
  dplyr::select(atencion, servicio, servicio_detalle, nivelPrioridad,salida_taquilla, cola_triage, 
                dur_triage, cola_servicio, dur_servicio) %>%
  filter_if(is.numeric, all_vars(. > 10)) %>% # Validado con asistenciales
  filter(cola_triage <= 14400, # Hasta 4 horas, validado con doctores de urgencias 
         cola_servicio <= 43200) # Hasta 24 horas, validado con asistenciales
```

## A tidy version of the clean data

```{r}
tidy_tiempos <- tiempos %>% 
  pivot_longer(cols = 5:8,
               names_to = "clase",
               values_to = "valor")

tidy_tiempos %>% 
  head()
```

### Considerations

- All the durations considered will last at least 10 seconds, based on medical expert criteria from the clinic.
- Triage queues higher than 4 hours will be discarded, the same goes for service time queues higher than 24 hours, based on medical expert criteria.

# Correlations and density plots

```{r}
tiempos %>%
  GGally::ggpairs(mapping = aes(alpha = 0.6), columns = 4:7)
```

All the event lengths seem to be uncorrelated with each other. At first glance some exponential distributions may appear.

# Descriptive statistics

```{r}
tiempos %>% 
  summary()
```

## Histograms

### Tiempo entre llegadas

```{r}
tibble(tiempo_llegadas) %>%
  ggplot(aes(x = tiempo_llegadas)) +
  geom_histogram(color = "black", fill = "skyblue")
```


Generales

```{r}
triaje <- tidy_tiempos %>%
  filter(clase == "cola_triage" | clase == "dur_triage") 

triaje %>% 
  ggplot(aes(x = valor, fill = clase)) +
  geom_histogram(color = "black", bins = 60) +
  facet_wrap(~clase, scales = "free", nrow = 2) +
  scale_x_continuous(labels = scales::comma) +
  scale_fill_brewer(palette = "Dark2") +
    labs(title = "Triaje",
     subtitle = "Distribuciones de los tiempos de cola y duraciones",
     x = "Duración (s)")

triaje %>% 
  filter(clase == "cola_triage") %>% 
  summarize(media = mean(valor), varianza = var(valor), sd = sd(valor), min = min(valor), max = max(valor))
  
```

### SAI Adultos

```{r}
sai <- tidy_tiempos %>% 
  filter(servicio_detalle == "Camillas SAI Adultos",
         clase == "cola_servicio" | clase == "dur_servicio")

sai %>% 
  ggplot(aes(x = valor, fill = clase)) +
  geom_histogram(color = "black", bins = 35) +
  facet_wrap(~servicio_detalle + clase, scales = "free_x", ncol = 2) +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "SAI Adultos",
    subtitle = "Distribuciones de los tiempos de cola y duraciones en los servicios",
    x = "Duración (s)")

sai %>%
  filter(clase == "dur_servicio") %>% 
  summarize(media = mean(valor), varianza = var(valor), sd = sd(valor), min = min(valor), max = max(valor))
```

### Trauma

```{r}
trauma <- tidy_tiempos %>% 
  filter(servicio_detalle == "Traumas Urgencias",
         #valor < 50000,
         #nivelPrioridad == "TRIAJE 5",
         #clase == "cola_servicio" | 
           clase == "dur_servicio") 

trauma %>% 
  ggplot(aes(x = valor, fill = clase)) +
  geom_histogram(color = "black", bins = 35) +
  facet_wrap(~servicio_detalle + clase, scales = "free_x", ncol = 2) +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "Traumas de urgencias",
    subtitle = "Distribuciones de los tiempos de cola y duraciones en los servicios",
    x = "Duración (s)")

trauma %>%
  filter(clase == "dur_servicio") %>% 
  summarize(media = mean(valor), varianza = var(valor), sd = sd(valor), min = min(valor), max = max(valor))
```

### Consultorio adultos


```{r}
adultos <- tidy_tiempos %>% 
  filter(servicio == "Urgencias",
         servicio_detalle != "Traumas Urgencias",
         servicio_detalle != "Camillas SAI Adultos",
         clase == "cola_servicio" | clase == "dur_servicio")

adultos %>% 
  ggplot(aes(x = valor, fill = clase)) +
  geom_histogram(color = "black", bins = 35) +
  facet_wrap(~servicio + clase, scales = "free_x", ncol = 2) +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "Consultorios de urgencias adultos",
    subtitle = "Distribuciones de los tiempos de cola y duraciones en los servicios",
    x = "Duración (s)")

adultos %>%
  filter(clase == "dur_servicio") %>% 
  summarize(media = mean(valor), varianza = var(valor), sd = sd(valor), min = min(valor), max = max(valor))
```

### Pediatría

```{r}
pediatria <- tidy_tiempos %>% 
  filter(servicio_detalle == "ERA Pediatria",
         clase == "cola_servicio" | clase == "dur_servicio") 

pediatria %>% 
  ggplot(aes(x = valor, fill = clase)) +
  geom_histogram(color = "black", bins = 35) +
  facet_wrap(~servicio + clase, scales = "free_x", ncol = 2) +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "Consultorios de urgencias Pediatría",
    subtitle = "Distribuciones de los tiempos de cola y duraciones en los servicios",
    x = "Duración (s)")

pediatria %>%
  filter(clase == "dur_servicio") %>% 
  summarize(media = mean(valor), varianza = var(valor), sd = sd(valor), min = min(valor), max = max(valor))
```

Distributions will only be fitted to the duration of the services, as the queues are of no interest, but rather an output of the discrete event simulation.

## Priority levels

```{r, fig.height=8, fig.width=9}
priority_levels <- tiempos %>%
  filter(servicio_detalle %in% c("Camillas SAI Adultos", "Traumas Urgencias", "ERA Pediatria", "Camillas SAI Pediatria", "Piso 4  Pediatria", "Bienestar Cirug\xeda", "Camillas Sala 1  Urgencias", "Camillas Sala 2  Urgencias", "Camillas Sala 3  Urgencias",
  "Camillas Sala 4  Urgencias", "Camillas Sala 5 Urgencias",  "Camillas Sala 6  Urgencias", "Camillas Sala 7  Urgencias",
  "Camillas Sala 8  Urgencias", "Fast track 1", "Fast track 2", "Fast track 3", "Sala de Reanimaci\xf3n", "Sala VIP UdeA")) %>% 
  group_by(nivelPrioridad) %>% 
  summarize(n = n(), 
            SAI = sum(servicio_detalle == "Camillas SAI Adultos", na.rm = TRUE)/n(),
            trauma = sum(servicio_detalle == "Traumas Urgencias", na.rm = TRUE)/n(),
            pediatria = sum(servicio_detalle == "ERA Pediatria" | servicio_detalle == "Camillas SAI Pediatria" | servicio_detalle == "Piso 4  Pediatria", na.rm = TRUE)/n(),
            adultos = sum(servicio == "Urgencias" & servicio_detalle != "Traumas Urgencias" & servicio_detalle != "Camillas SAI Adultos" & servicio_detalle != "Camillas SAI Pediatria" & servicio_detalle != "Piso 4  Pediatria", na.rm = TRUE)/n()) 

priority_levels

priority_levels %>% 
  pivot_longer(cols = 3:6) %>%
  mutate(name = as_factor(name),
         nivelPrioridad = fct_relevel(nivelPrioridad, "TRIAJE 1", "TRIAJE 2", "TRIAJE 3", "TRIAJE 4")) %>% 
  ggplot(aes(x = name, y = value, fill = name)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~nivelPrioridad, ncol = 1) +
  theme_economist() +
  scale_fill_economist() +
  labs(x = NULL,
       y = "Probabilidades para cada triaje de ir a un servicio",
       fill = "Servicio")
```

## probabilidad de triaje

```{r, fig.height=8, fig.width=9}
priority_triaje <- tiempos %>%
  filter(servicio_detalle %in% c("Camillas SAI Adultos", "Traumas Urgencias", "ERA Pediatria", "Camillas SAI Pediatria", "Piso 4  Pediatria", "Bienestar Cirug\xeda", "Camillas Sala 1  Urgencias", "Camillas Sala 2  Urgencias", "Camillas Sala 3  Urgencias",
  "Camillas Sala 4  Urgencias", "Camillas Sala 5 Urgencias",  "Camillas Sala 6  Urgencias", "Camillas Sala 7  Urgencias",
  "Camillas Sala 8  Urgencias", "Fast track 1", "Fast track 2", "Fast track 3", "Sala de Reanimaci\xf3n", "Sala VIP UdeA")) %>% 
  group_by(nivelPrioridad) %>% 
  summarize(n = n(), 
            SAI = sum(servicio_detalle == "Camillas SAI Adultos", na.rm = TRUE)/n(),
            trauma = sum(servicio_detalle == "Traumas Urgencias", na.rm = TRUE)/n(),
            pediatria = sum(servicio_detalle == "ERA Pediatria" | servicio_detalle == "Camillas SAI Pediatria" | servicio_detalle == "Piso 4  Pediatria", na.rm = TRUE)/n(),
            adultos = sum(servicio == "Urgencias" & servicio_detalle != "Traumas Urgencias" & servicio_detalle != "Camillas SAI Adultos" & servicio_detalle != "Camillas SAI Pediatria" & servicio_detalle != "Piso 4  Pediatria", na.rm = TRUE)/n()) 

priority_levels

priority_levels %>% 
  pivot_longer(cols = 3:6) %>%
  mutate(name = as_factor(name),
         nivelPrioridad = fct_relevel(nivelPrioridad, "TRIAJE 1", "TRIAJE 2", "TRIAJE 3", "TRIAJE 4")) %>% 
  ggplot(aes(x = name, y = value, fill = name)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~nivelPrioridad, ncol = 1) +
  theme_economist() +
  scale_fill_economist() +
  labs(x = NULL,
       y = "Probabilidades para cada triaje de ir a un servicio",
       fill = "Servicio")
```


# Distribution fitting and hypothesis testing

In this part, several distribution will be tried for every service, based on the recommendations obtained with the Cullen and Frey method, further reference can be obtained in: Cullen AC and Frey HC (1999), Probabilistic techniques in exposure assessment. Plenum Press, USA, pp. 81-159.

## Duración Triaje

### Exponential distribution

```{r}
fit_triage_exp <- fitdist(tiempos$dur_triage, distr = "exp", method = "mle")
fit_triage_exp
plot(fit_triage_exp)
```

Null hyphotesis rejected based on the Q-Q plot, the data can't possibly be represented by an exponential distribution.

### Gamma distribution

```{r}
fit_triage_gamma <- fitdist(tiempos$dur_triage, distr = "gamma", method = "mle")
fit_triage_gamma
plot(fit_triage_gamma)
```

Null hyphotesis could be rejected, need further testing.

### Lognormal

```{r}
fit_triage_lognormal <- fitdist(tiempos$dur_triage, distr = "lnorm", method = "mle")
fit_triage_lognormal
plot(fit_triage_lognormal)
```

Data is definitely lognormal, need to conduct formal hypothesis testing here.

## Camillas SAI Adultos

```{r}
tiempos_SAI <- tiempos %>% 
  filter(servicio_detalle == "Camillas SAI Adultos")

descdist(as.numeric(tiempos_SAI$dur_servicio), boot = 500)
```

### Exponential

```{r}
fit_SAI_exp <- fitdist(tiempos_SAI$dur_servicio/1000, distr = "exp", method = "mle")
fit_SAI_exp
plot(fit_SAI_exp)
```

The fit doesn't work anyway, so no need to think about it anyway, because the PDF will be the same for a scaled version.

### Lognormal

```{r}
fit_SAI_lognormal <- fitdist(tiempos_SAI$dur_servicio, distr = "lnorm", method = "mle")
fit_SAI_lognormal
plot(fit_SAI_lognormal)
```

### Gamma distribution

Need an scaling factor for the gamma to work, **need to reference this.**

```{r}
fit_SAI_gamma <- fitdist(tiempos_SAI$dur_servicio, distr = "gamma", method = "mle", lower=c(0,0))
fit_SAI_gamma
plot(fit_SAI_gamma)
```

SAI seems to be Gamma!

## Traumas Urgencias

```{r}
tiempos_trauma <- tiempos %>% 
  filter(servicio_detalle == "Traumas Urgencias" 
         & dur_servicio < 100000
         )

#descdist(as.numeric(tiempos_trauma$dur_servicio), boot = 500)
```

### Exponential

```{r}
fit_trauma_exp <- fitdist(tiempos_trauma$dur_servicio/1000, distr = "exp", method = "mle")
fit_trauma_exp
plot(fit_trauma_exp)
```

The fit doesn't work anyway, so no need to think about it anyway, because the PDF will be the same for a scaled version.

### Lognormal

```{r}
fit_trauma_lognormal <- fitdist(tiempos_trauma$dur_servicio, distr = "lnorm", method = "mle")
fit_trauma_lognormal
plot(fit_trauma_lognormal)
```

trauma seems to be lognormal.

### Gamma distribution

```{r}
fit_trauma_gamma <- fitdist(tiempos_trauma$dur_servicio, distr = "gamma", method = "mle", lower=c(0,0))
fit_trauma_gamma
plot(fit_trauma_gamma)
```

No.

## ERA Pediatria

```{r}
tiempos_pediatria <- tiempos %>% 
  filter(servicio_detalle == "ERA Pediatria")

descdist(as.numeric(tiempos_pediatria$dur_servicio), boot = 500)
```

### Exponential

```{r}
fit_pediatria_exp <- fitdist(tiempos_pediatria$dur_servicio/1000, distr = "exp", method = "mle")
fit_pediatria_exp
plot(fit_pediatria_exp)
```

The fit doesn't work anyway, so no need to think about it anyway, because the PDF will be the same for a scaled version.

### Lognormal

```{r}
fit_pediatria_lognormal <- fitdist(tiempos_pediatria$dur_servicio, distr = "lnorm", method = "mle")
fit_pediatria_lognormal
plot(fit_pediatria_lognormal)
```

Parece que sí

### Gamma distribution

```{r}
fit_pediatria_gamma <- fitdist(tiempos_pediatria$dur_servicio, distr = "gamma", method = "mle", lower=c(0,0))
fit_pediatria_gamma
plot(fit_pediatria_gamma)
```
 
No.

In case a triangular distribution was needed you could do:

`fit_pediatria_triangular <- fitdist(tiempos_pediatria$dur_servicio, distr = "triang", method = "mge",  start = list(min=0, mode=25000,max=100000), gof = "CvM")`


## Consultorios adultos

```{r}
tiempos_adultos <- tiempos %>% 
  filter(servicio == "Urgencias",
         servicio_detalle != "Traumas Urgencias",
         servicio_detalle != "Camillas SAI Adultos")

descdist(as.numeric(tiempos_adultos$dur_servicio), boot = 500)
```

### Exponential

Can I use a scaling factor and after that divide the rate parameter also?

```{r}
fit_adultos_exp <- fitdist(tiempos_adultos$dur_servicio/1000, distr = "exp", method = "mle")
fit_adultos_exp
plot(fit_adultos_exp)
```

The fit doesn't work anyway, so no need to think about it anyway, because the PDF will be the same for a scaled version.

### Lognormal

```{r}
fit_adultos_lognormal <- fitdist(tiempos_adultos$dur_servicio, distr = "lnorm", method = "mle")
fit_adultos_lognormal
plot(fit_adultos_lognormal)
```

Parece que sí

### Gamma distribution

```{r}
fit_adultos_gamma <- fitdist(tiempos_adultos$dur_servicio, distr = "gamma", method = "mle", lower=c(0,0))
fit_adultos_gamma
plot(fit_adultos_gamma)
```

## Tiempo llegadas

```{r}
cullen_fray_llegadas <- descdist(as.numeric(tiempo_llegadas), boot = 1000)
cullen_fray_llegadas
```


### Weibull distribution

```{r}
fit_llegadas_weibull <- fitdist(as.numeric(tiempo_llegadas), distr = "weibull", lower=c(0,0))
fit_llegadas_weibull
plot(fit_llegadas_weibull)
```

This will be the selected distribution for this variable. May need to consider the kind of triage to correctly estimate a distribution.


### Gamma distribution

```{r}
fit_llegadas_gamma <- fitdist(as.numeric(tiempo_llegadas), distr = "gamma", method = "mle", lower=c(0,0))
fit_llegadas_gamma
plot(fit_llegadas_gamma)
```

Null hyphotesis could be rejected, need further testing.

### Lognormal

```{r}
fit_llegadas_lognormal <- fitdist(as.numeric(tiempo_llegadas), distr = "lnorm", method = "mle")
fit_llegadas_lognormal
plot(fit_llegadas_lognormal)
```


### Exponential distribution

```{r}
fit_llegadas_exp <- fitdist(as.numeric(tiempo_llegadas/10), distr = "exp")
fit_llegadas_exp
plot(fit_llegadas_exp)
```


```{r, include=FALSE}
write_csv(tiempos, "datasets/tiempos.csv")
write_lines(tiempo_llegadas, "datasets/tiempo_llegadas.txt")
```

